---
title: "Sea Turtle Database Exploration"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(googlesheets4)
library(dplyr)
library(tidyr)
library(lubridate)
library(plotly)

# Authenticate without prompting (uses GOOGLE_SERVICE_ACCOUNT_JSON secret in CI,
# or your cached token locally — run gs4_auth() once on your machine first)

if (Sys.getenv("GOOGLE_SERVICE_ACCOUNT_JSON") != "") {
  gs4_auth(path = Sys.getenv("GOOGLE_SERVICE_ACCOUNT_JSON"))
} else {
  gs4_auth()  # uses cached token locally
}

sheet_id <- "1zOO7FV_caqNCEn9OUmxF5kKYrCUEO0liPwRCGqR7QnE"
```

# Load the three sheets

```{r load}
db <- read_sheet(sheet_id, sheet = "Sea Turtle Database") %>% 
   rename(date = `Activity Date YYYY-MM-DD`,
         species = Species) %>%
  mutate(date = ymd(date))

effort <- read_sheet(sheet_id, sheet = "Effort", col_types="c") %>%
  rename(date = `patrol date YYYY-MM-DD`) %>%
  mutate(date = ymd(date))

effort$time_on_patrol_mins <- (as.numeric(substr(effort$`time on patrol HH:MM:SS`,1,2))*60) + as.numeric(substr(effort$`time on patrol HH:MM:SS`,4,5))

hatchery <- read_sheet(sheet_id, sheet = "Hatchery") %>%
  rename(date = `Relocation Date YYYY-MM-DD`,
         species = Species) %>%
  mutate(date = ymd(date))
```

# Observations through time

Each point is one record (database row) on the date it was logged. Useful for spotting gaps and overlaps in coverage across the three sources.

```{r through-time, fig.height=7}
obs_long <- bind_rows(
  db       %>% transmute(date, sheet = "Sea Turtle Database"),
  effort   %>% transmute(date, sheet = "Effort"),
  hatchery %>% transmute(date, sheet = "Hatchery")
) %>%
  filter(!is.na(date)) %>%
  count(sheet, date, name = "n")

obs_long$sheet <- factor(obs_long$sheet,
                         levels = c("Sea Turtle Database", "Effort", "Hatchery"))

p_list <- lapply(levels(obs_long$sheet), function(s) {
  plot_ly(obs_long %>% filter(sheet == s),
          x = ~date, y = ~n,
          type = "scatter", mode = "markers",
          marker = list(size = 5, opacity = 0.6),
          name = s,
          hovertemplate = paste0(s, "<br>%{x|%Y-%m-%d}<br>%{y} records<extra></extra>")) %>%
    layout(yaxis = list(title = s))
})

subplot(p_list, nrows = 3, shareX = TRUE, titleY = TRUE) %>%
  layout(title = "Records through time — by sheet",
         showlegend = FALSE)
```

# Total records (database rows) per year — all three sheets together

Click any line in the legend to toggle it on/off. This is the view for checking when effort and records line up.

```{r yearly-totals}
yearly <- bind_rows(
  db       %>% transmute(date, sheet = "Sea Turtle Database"),
  effort   %>% transmute(date, sheet = "Effort"),
  hatchery %>% transmute(date, sheet = "Hatchery")
) %>%
  filter(!is.na(date)) %>%
  mutate(year = year(date)) %>%
  count(sheet, year, name = "records")

plot_ly(yearly,
        x = ~year, y = ~records,
        color = ~sheet, colors = "Set1",
        type = "scatter", mode = "lines+markers",
        hovertemplate = "%{x}<br>%{y} records<extra>%{fullData.name}</extra>") %>%
  layout(
    title = "Total records per year by sheet",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Records"),
    legend = list(title = list(text = "<b>Click to toggle</b>"))
  )
```

# Species detected per survey night through time

Counts of nesting/sighting records per night, coloured by species. Toggle species in the legend.

```{r species-per-night}
db_factor <- db
db_factor$date <- as.factor(db_factor$date)
db_factor$species <- as.factor(db_factor$species)

species_nightly <- db_factor %>%
  filter(!is.na(date), !is.na(species), species != "") %>%
  count(date, species, name = "records", .drop = FALSE)

species_nightly$date <- ymd(species_nightly$date)

plot_ly(species_nightly,
        x = ~date, y = ~records,
        color = ~species, colors = "Dark2",
        mode = "lines",
        marker = list(size = 4, opacity = 0.7),
        hovertemplate = "%{x|%Y-%m-%d}<br>%{y} records<extra>%{fullData.name}</extra>") %>%
  layout(
    title = "Records per survey night",
    xaxis = list(title = ""),
    yaxis = list(title = "Records per night"),
    legend = list(title = list(text = "<b>Species</b>"))
  )
```

# Effort through time

Using the start and end times in the database we can look at person hours per day.

```{r effort}
effort_factor <- effort
effort_factor$date <- as.factor(effort_factor$date)
daily_effort <- effort_factor %>%
  group_by(date, .drop = FALSE) %>%
  summarise(total_effort = sum(time_on_patrol_mins))

daily_effort$date <- ymd(daily_effort$date)

plot_ly(daily_effort,
        x = ~date, y = ~total_effort,
        mode = "lines",
        marker = list(size = 4, opacity = 0.7),
        hovertemplate = "%{x|%Y-%m-%d}<br>%{y} minutes<extra></extra>") %>%
  layout(
    title = "Survey duration per day",
    xaxis = list(title = ""),
    yaxis = list(title = "Minutes")
  )
```

# Records per unit of time (minutes)

```{r encounter-rate}
tmp <- db[!is.na(db$species), ] %>% group_by(date) %>% summarise(n())

encounter_rate <- left_join(daily_effort, tmp)
encounter_rate$`n()`[is.na(encounter_rate$`n()`)] <- 0
encounter_rate$rate_hr <- encounter_rate$`n()` / (encounter_rate$total_effort / 60)

encounter_rate <- encounter_rate[!is.na(encounter_rate$rate_hr), ]
encounter_rate <- encounter_rate[encounter_rate$rate_hr < 1000, ]

plot_ly(encounter_rate,
        x = ~date, y = ~rate_hr,
        mode = "lines",
        marker = list(size = 4, opacity = 0.7),
        hovertemplate = "%{x|%Y-%m-%d}<br>%{y} encounters per hr<extra></extra>") %>%
  layout(
    title = "Records per hour",
    xaxis = list(title = ""),
    yaxis = list(title = "Records per hour of survey")
  )
```

Some of those look wildly implausible!

```{r encounter-hist}
plot_ly(encounter_rate, y = ~rate_hr, type = "histogram")
```

Which dates have the really high rates?

```{r high-rates}
tmp <- encounter_rate[encounter_rate$rate_hr > 20, ]
tmp[order(tmp$rate_hr, decreasing = TRUE), ]
```
